---
title: "Locality sensitive hashing & Recommender systems"
author: "Hugo Trivino"
output: 
  html_notebook:
    toc: yes
    toc_float: yes
---

```{r}
setwd(".")
set.seed(1122)
options("digits"=3)
library(textreuse)
library(lsa)

```

#2.1 Topic: Locality sensitive hashing

##(a) Visualizing our dataset as a characteristic matrix, what is the size (rows and columns) of our characteristic matrix?


```{r}
files <- list.files("./hw4.movies",full.names = T)
corpus <-TextReuseCorpus(files, tokenizer = tokenize_ngrams, n = 5,
keep_tokens = TRUE,progress = F)
shingles <-c()
for (x in 1:length(corpus)){
  h <- paste0("000",x)
  user <-"user"
  user <- paste0(user,substr(h,start=nchar(h)-2,nchar(h)))
  shingles <- union(shingles,tokens(corpus[[user]]))
}
cat(
  paste0("(a) The characteristic matrix has ",length(shingles), " rows and ", length(corpus)," Columns\n")
)
rm(h,shingles,x,user)
```
##(b) Let’s take a look at user ID 20. For this user, answer the following questions by examining the object returned from corpus[["user20"]]

```{r}
d <- corpus[["user020"]]
cat(
  paste0("(i)\tThis user has rated: ",stringr::str_count(d$content,"\n")," movies\n")
)
cat("(ii)\tThe first five shingles for this user are:\n")
for (x in 1:5){
  cat(paste("\t",x,"-> ",tokens(d)[x],"\n"))
}
rm(x,d)
```


##(c)Let’s find similar users using the brute-force pairwise method first. Using pairwise_candidates() and pairwise_compare(), to get a tibble that contains all of the pairwise candidates.
```{r}
res <- pairwise_candidates(pairwise_compare(corpus, jaccard_similarity,progress = F))

```



```{r}
ans=""
for (x in which(res$score>=0.4)){
  ans<-paste0(ans,res[x,1]," <--> ")
  ans<-paste0(ans,res[x,2]," score of -> ")
  ans<-paste0(ans,res[x,3],"\n\t ")
  
}
cat(
  paste(
    " (i)\tThere are ",length(which(res$score>=0.6))," pairs with score at least 0.60\n",
    "(ii)\tThere are ",length(which(res$score>=0.5))," pairs with score at least 0.50\n",
    "(iii)\tThere are ",length(which(res$score>=0.4))," pairs with score at least 0.40\n",
    "(iv)\tThe pairs with a score of at least 0.4 are:\n\t",
    ans
  )
)
rm(ans,x)
```
##(d)Now, let’s prepare to use LSH and find candidate pairs in a more optimized manner. First, we determine the parameters (number of minhashes and bands).
```{r}

#print("(iv) The pairs with a score of at least 0.4 are:")
r=1
b=5

cat(paste0(
  "(i) With ",r*b," minhash signatures and ",
  b, " bands, we can obtain a LSH probability of ",
  lsh_probability(h=r*b,  b=b, s=0.60),"\n"
  #format(lsh_probability(h=r*b,  b=b, s=0.60)),digits=5)
  )
)
minhash <- minhash_generator(n=r*b, seed=100)
corpus2 <- TextReuseCorpus(files, tokenizer = tokenize_ngrams, n = 5,
                          minhash_func = minhash, keep_tokens = TRUE,progress = F)
cat(
  "(ii) The first minhashes of user020 are:\n\t",
  paste(minhashes(corpus2[["user020"]]),"\n\t")
)

```
##(e) Now run LSH and find candidate pairs in an optimized way. Then answer the following:
```{r}
lsh_threshold(h=b,b=5)
buckets <- lsh(corpus2, bands = 5, progress=F)
candidates2 <- lsh_candidates(buckets)
res2 <- lsh_compare(candidates2, corpus, jaccard_similarity,progress = F)
```

```{r}

ans=""
for (x in which(res2$score>=0.4)){
  ans<-paste0(ans,res2[x,1]," <--> ")
  ans<-paste0(ans,res2[x,2]," score of -> ")
  ans<-paste0(ans,res2[x,3],"\n\t ")
  
}
cat(
  paste0(
    "(i) I expect to find zero candidates with a Jaccard similarity between 0.60 and 0.50",
         " and there are ",length(which(res2$score>=0.5)), " candidate pairs.",
         "I do not expect any candidate pairs since I expect to find 80% of s=0.6. And there are not pairs with such Jaccard similarity or with s >=0.5.\n",
    "(ii)\t The pairs with a score of at least 0.4 are:\n\t ",
    ans,
    "\n  There were only ", length(res2$a), " comparisons.\n",
    "(iii) Only 75% are the same as shown in part (c)(iv).\n",
    "(iv) The number of comparisons performed in the optimized version is ",
    format(length(res2$a)/length(res$a)*100,digits = 4), "% of the original comparison. ",
    "Since there were only ",length(res2$a)," comparisons instead of ",length(res$a)," we saved " , format(100-length(res2$a)/length(res$a)*100,digits = 4),"% of the work.\n"
  )
)

rm(list=ls())

```

#2.2 Topic: Content-based recommendation system (2 Points)
```{r}

ratings <-  read.csv(file = "ml-latest-small/ratings.csv",sep = ",")

movies <-read.csv(file = "ml-latest-small/movies.csv",sep = ",")
genreVector <-c()

for (genreLst in movies[,3]){
  genxmovie<-strsplit(genreLst,split = "|",fixed = T)
  for (genre in genxmovie){
    genreVector<-union(genre,genreVector)
  }
  if(length(genreVector)>=20){
    break
  }
}
genreVector <- sort(genreVector)
createUserProfile <-function(userID){
    moviesId <-c(ratings[which(ratings$userId== userID),c(2)])
    genreMatrix<-c()
    moviesIdx <- which(movies$movieId %in% moviesId)
    for (genreLst in movies[moviesIdx,3]){
      genxmovie<-strsplit(genreLst,split = "|",fixed = T)
      for (genre in genxmovie){
        genreMatrix<-append(genreMatrix,genre)
      }
    }
    userProfile <-c()
    for (genre in genreVector){
      userProfile <-append(userProfile,length(which(genreMatrix==genre)))
    }
    userProfile <- userProfile /length(moviesIdx)
    return(userProfile)
}
createMovieProfile <-function(movieId){
  movieIdx <- which(movies$movieId == movieId)
  genres <- as.character(movies[movieIdx,3]) 
  genres <- strsplit(genres,"|",fixed=T)
  vG<-c()
  for (g in genres){
    vG <- union(vG,g)
  }
  genres <-vG
  genres <-as.factor(genres)
  movieProfile <- c()
  for (genre in genreVector){
    movieProfile<-append(movieProfile,genre %in% genres *1)
  }
  return(movieProfile)
}

set.seed(1122)
user<- 20391413 %% 671
#user<-1
uservector <-createUserProfile(user)
moviesSeen <- ratings$movieId[which(ratings$userId==user)]
moviestorecomend <-sample(movies$movieId,size = length(moviesSeen)+10)
moviestorecomend <-sample(setdiff(moviestorecomend,moviesSeen),size = 10)
recommendable <-c()
movieName<-c()
for (movie in moviestorecomend){
  recommendable <- append(recommendable,cosine(createMovieProfile(movie),uservector))
  movieName <- rbind(movieName, movies[which(movies$movieId==movie),] )
}

recommender <-data.frame(movieName[,1:2],recommendable,row.names = 1:10)
recommender <- data.frame(recommender[order(recommender$recommendable,decreasing = T),],row.names = 1:10)
print(recommender[1:5,])

```

#2.3  Topic: Collaborative Filtering (2 points)
##(a) Prediction using user-user similarity:
```{r}
user <- 191
userRating <-ratings[which(ratings$userId ==user),]
actualRating <-userRating[which(userRating$movieId %in% c(150,296,380,590)),]
userRating[which(userRating$movieId %in% c(150,296,380,590)),]$rating <- NA
#print(userRating)

otherUsers <-c(375, 225, 513, 50, 657)
simUser <- c(0.3049645,0.2420382,0.4358974,0.2009804, 0.2262774)
rtnOther <-c()
rtnMatrix<-data.frame()
for (other in otherUsers){
  rtnOther <-c()
  otherUserRating<-ratings[which(ratings$userId==other),]
  for (movieToRate in c(150,296,380,590)){
    if (length(which(otherUserRating$movieId==movieToRate))>0){
      rtnOther<-cbind(rtnOther,otherUserRating[which(otherUserRating$movieId==movieToRate),]$rating)
    }else{
      rtnOther<-cbind(rtnOther,NA)
    }
  }
  rtnMatrix <-rbind(rtnMatrix,rtnOther)
  #print(otherUserRating[which(otherUserRating$movieId %in% c(150,293,380,590)),])
}

colnames(rtnMatrix)<-c(150,296,380,590)
row.names(rtnMatrix)<-otherUsers
finalRating <-c(0,0,0,0)
for (x in 1:3){
  for (y in 1:4){
    finalRating[y] <-finalRating[y]+rtnMatrix[x,y] * simUser[x]/sum(simUser[1:3])
  }
  
}
rmse <-0
cat(paste0("User ID 191, 5 random user IDs: 375, 225, 513, 50, 657\n",
           "Using user-user similarity, User ID 191 will rate the movies as follows:\n"
           ))


for (c in 1:4) {
  cat(paste0(c(150,296,380,590)[c],":",format(finalRating[c],digits=4),"\n"))
  rmse <- 1/4*((actualRating$rating[c] - finalRating[c])^2)^0.5 + rmse
}
cat(paste0("RMSE:",format(rmse,digits=4),"\n"))
```
##(b) Prediction using item-item similarity:
```{r}
user <- 191
userRating <-ratings[which(ratings$userId ==user),]
actualRating <-userRating[which(userRating$movieId %in% c(150,296,380,590)),]
userRating[which(userRating$movieId %in% c(150,296,380,590)),]$rating <- NA
#print(userRating)
moviesSim <-c()
moviessimMatrix <-c()
moviesWatched <- setdiff(userRating$movieId,c(150,296,380,590))
for (x in c(150,296,380,590)){
  profileMovie <- createMovieProfile(x)
  for (m in moviesWatched){
    moviesSim <- append(moviesSim,cosine(createMovieProfile(m),profileMovie))
  }
  moviessimMatrix <-cbind(moviessimMatrix,moviesSim)
  moviesSim <-c()
}
#moviessimMatrix <- data.frame(moviesSim[1:length(moviesWatched)],moviesSim[length(moviesWatched)+1:2*length(moviesWatched)],)
moviessimMatrix<-cbind(moviessimMatrix,moviesWatched)

finalRating <-c(0,0,0,0)
for (m in 1:4){
  neighboor <- moviessimMatrix[order(moviessimMatrix[,m],decreasing=T),]
  for (r in 1:3){
    finalRating[m] <- neighboor[r,m]/sum(neighboor[1:3,m])*userRating$rating[userRating$movieId==neighboor[r,5]] + finalRating[m] 
}
}
cat(paste0("User ID 191, 5 random user IDs: 375, 225, 513, 50, 657\n",
           "Using item-item similarity, User ID 191 will rate the movies as follows:\n"
           ))


for (c in 1:4) {
  cat(paste0(c(150,296,380,590)[c],":",format(finalRating[c],digits=4),"\n"))
  rmse <- 1/4*((actualRating$rating[c] - finalRating[c])^2)^0.5 + rmse
}
cat(paste0("RMSE:",format(rmse,digits=4),"\n"))


```


